---
title: "Final Report Index"
format: html
execute:
  wrap: true
  output-dir: docs
---

## Introduction

Relational databases rely on well-defined structures to reduce repetition, ensure consistency, and support efficient data usage. In practice, however, data analysts work with data sets that are messy, with unclear keys. While there are existing R packages, such as testdat, that offer tools for data quality, support for functions that directly assess normal forms or assist in finding keys is limited. Our project addresses that gap by creating an R package that provides intuitive functionality for determining keys and evaluating normal forms. Our package, called NormForm, introduces a set of tools to help R users identify potential keys, test whether selected attributes are valid keys, and determine whether or not a data set satisfies the first and second normal form requirements. Collectively, these functions support more efficient data organization and improve the reliability of analyses, along with providing a structured approach to preparing data sets for relational analysis. The NormForm bridges the gap between theoretical principles and practical application, making relational database concepts more accessible to all users.

## Motivation

The motivation for developing this package stems from previous issues encountered while working with data sets that lacked clear key structures or violated normalization principles. Determining if a data set meets first or second normal form requirements often requires manual inspection, and identifying valid keys can become computationally expensive as the number of variables increases. Our goal was to develop a set of tools that would make these tasks more accessible to students, researchers, and anyone working with R. In order to accomplish this, we developed four functions. Additionally, we used three data sets to serve as examples for how the functions work. For each function the logic is outlined, and working drafts of each function were created and tested on several sample data sets:

-   college: 1372 observations, 29 variables, unique ID variable (unitid), contains data over U.S. colleges and student demographics

-   auto: 398 observations, 11 variables, no unique ID variable, contains vehicle build and performance data

-   counties: 100 observations, 18 variables, unique ID variable (RN), contains U.S. census data from 1994

```{r}
data("auto")
data("college")
data("counties")
```

### Functions

#### is_key

```{r, warning=FALSE}
library(dplyr)
library(purrr)
library(assertthat)

is_key <- function(data, cols) {
  assert_that(
    is.data.frame(data),
    not_empty(data)
  )
  !any(duplicated(data[, cols, drop = FALSE]))
}
```

#### find_keys

```{r}
find_keys <- function(data, max_size = 2) {
  cols <- colnames(data)
  n <- length(cols)
  all_keys <- list()
  
  for (k in 1:min(max_size, n)) {
    combos <- combn(cols, k, simplify = FALSE)
    
    for (combo in combos) {
      if (is_key(data, all_of(combo))) {
        all_keys <- append(all_keys, list(combo))
      }
    }
  }
  
  minimal_keys <- list()
  
  for (key in all_keys) {
    is_minimal <- TRUE
    
    for (other in all_keys) {
      if (length(other) < length(key) &&
          all(other %in% key)) {
        is_minimal <- FALSE
        break
      }
    }
    
    if (is_minimal) {
      minimal_keys <- append(minimal_keys, list(key))
    }
  }
  
  if (length(minimal_keys) == 0) {
    return(NULL)
  }
  
  tibble(key_size = sapply(minimal_keys, length), 
         columns = sapply(minimal_keys, paste, collapse = ", ")
  )
}
```

#### first_normal

```{r}
first_normal <- function(data, max_size = 2) {
  
  is_atomic <- function(x) {
    all(!sapply(x, function(v) length(v) != 1 || is.list(v)))
  }
  
  atomic_cols <- sapply(data, is_atomic)
  
  if (!all(atomic_cols)) {
    return("FALSE: Dataset is NOT in First Normal Form, non-atomic columns found")
  }

      result <- find_keys(data, max_size)

  if (is.null(result) | is.character(result)) {
    return("FALSE: Dataset is NOT in First Normal Form, no candidate key exists")
  }
  
  return("TRUE")
}
```

#### second_normal

```{r}
second_normal <- function(data, max_size = 2) {

  result <- first_normal(data, max_size = max_size)
  key_results <- find_keys(data, max_size = max_size)

  if (grepl("FALSE", result)) {
    return("FALSE: Dataset is NOT in Second Normal Form because it is not in First Normal Form")
  }

  for (i in seq_len(nrow(key_results))) {

    key_cols <- unlist(strsplit(key_results$columns[i], ", "))

    if (length(key_cols) > 1) { 
      
      subsets <- unlist(
        lapply(1:(length(key_cols) - 1),
               function(k) combn(key_cols, k, simplify = FALSE)
        ),
        recursive = FALSE
      )

      for (subset in subsets) {

        if (!any(duplicated(data[subset]))) {
          return(paste0(
            "FALSE: Dataset is NOT in Second Normal Form. ",
            "Partial dependency detected: subset {",
            paste(subset, collapse = ", "),
            "} uniquely identifies rows."
          ))
        }
      }
    }
  }

  return("TRUE")
}
```

#### third_normal

```{r}
third_normal <- function(data, max_size = 2) {
  
  result_2nf <- second_normal(data, max_size = max_size)
  if (grepl("FALSE", result_2nf)) {
    return("FALSE: Dataset is NOT in Third Normal Form because it is not in Second Normal Form")
  }
  
  key_results <- find_keys(data, max_size = max_size)
  
  if (is.null(key_results) || nrow(key_results) == 0) {
    return("FALSE: No candidate keys found")
  }
  
  all_cols <- colnames(data)
  
  keys_list <- strsplit(key_results$columns, ", ")
  prime_attrs <- unique(unlist(keys_list))
  non_key_attrs <- setdiff(all_cols, prime_attrs)

  if (length(non_key_attrs) <= 1) return("TRUE")

  all_subsets <- list()
  for (k in 1:min(2, length(non_key_attrs))) {
    all_subsets <- c(all_subsets, combn(non_key_attrs, k, simplify = FALSE))
  }

  for (target in non_key_attrs) {
    for (subset in all_subsets) {
      
      if (target %in% subset) next
      
      if (is_key_cached(data, subset)) next
      
      if (no_duplicates_cached(data, subset)) {
        return(paste0(
          "FALSE: Dataset is NOT in Third Normal Form. ",
          "Possible transitive dependency: {",
          paste(subset, collapse = ", "),
          "} -> ", target,
          ". (May be data-specific coincidence)"
        ))
      }
    }
  }
  
  return("TRUE")
}
```

#### Wrappers

```{r}
key_cache <- new.env(parent = emptyenv())

is_key_cached <- function(data, cols) {
  key <- paste(sort(cols), collapse = "|")
  
  if (exists(key, envir = key_cache)) {
    return(get(key, envir = key_cache))
  }
  
  val <- is_key(data, cols)
  assign(key, val, envir = key_cache)
  val
}

dup_cache <- new.env(parent = emptyenv())

no_duplicates_cached <- function(data, cols) {
  key <- paste(sort(cols), collapse = "|")
  
  if (exists(key, envir = dup_cache)) {
    return(get(key, envir = dup_cache))
  }
  
  val <- !any(duplicated(data[, cols, drop = FALSE]))
  assign(key, val, envir = dup_cache)
  val
}
```

### Descriptions

#### is_key

Description: This function checks whether a user-specified set of columns is a valid key. The two required arguments include the data set and a list of column(s) being inquired. Output returns a logical result after testing if the listed column(s) is a unique identifier of the data.

```{r}
print(is_key)
```

Example:

```{r}
is_key(college, c("region", "city"))
is_key(college, "unitid")
```

Here are two examples of how is_key can be used to check if a column or list of columns are a key. The FALSE output tells us region and city are not a combination key because the rows are not unique. Unitid is a unique identifier given the TRUE result, and therefore a key for the college data.

#### find_keys

Description: This function searches for the smallest possible keys in the data set and can be extended to longer lengths with argument max_size to control the size of combinations considered. Keys are found by cycling through subset combinations of columns and using is_key to test the key status. Output includes all keys that uniquely identify the data set that are the equal to or less than the specified max_size argument.

```{r}
print(find_keys)
```

Example:

```{r, warning=FALSE}
find_keys(counties, max_size = 1) 
find_keys(counties)
```

In this example, there are uniquely identifying keys of one column. The RN is a random assigned number to each county, but the other variables identified as a key would not make appropriate keys given that they are numerical observations of the county. In the case that there are not appropriate single-column keys, the optional max_size argument allows to look for larger, more robust keys.

Example

```{r}
find_keys(auto, max_size = 2)
find_keys(auto, max_size = 3)
```

Using the auto data, which has unique identifying variable, we can see that there are no keys of size 2 or smaller, resulting in an error message. By increasing the max_size, the function is able to identify keys of size 3.

#### first_normal

Description: This function returns a logical value indicating whether the data set is in first normal form. The first condition of first normal form is all atomic columns and then uses the find_key function to identify a key. If the data is tidy and has a key, then the function will return TRUE and key(s) have been found. The function will return FALSE and will return a message stating which condition is not met

```{r}
print(first_normal)
```

Example

```{r}
first_normal(college)

first_normal(auto, max_size = 2)
```

Using the first_normal function on the college data reveals that it is in first normal form by having atomic columns and at least one key. In contrast, the auto data set is not in first normal form since no keys of size 2 or smaller exist.

```{r}
first_normal(auto, max_size = 3)
```

When we raise the max_size of keys to 3, the auto data can now be considered to be in first normal form.

#### second_normal

Description: This function tests for second normal form conditions and returns a logical value like that of first_normal. The first condition is that the data much be in first normal form, so the function first_normal is used to verify. The second condition requires that no partial dependencies exists, meaning that non-key attribute depend on the entirity of the primary key. A key consisting of a single attribute is automatrically in second normal form. To check for partial dependencies for keys with \>1 columns, every possible subset combination will be created and tested for duplicates. If there are no duplicates, then the subset is a smaller unique key and therefore the original key has a partial dependency. The output of this function include a logical result, and a message stating which condition was not met when the result is FALSE.

```{r}
print(second_normal)
```

Example

```{r}
second_normal(college)
```

The function reveals that the college data set is in second normal form without needing any alteration. Notably, running only this function on a data set and receiving a TRUE result implies that at least one key exists and the data is in first normal form.

Example

```{r}
second_normal(auto, max_size = 2)
second_normal(auto, max_size = 3)
```

When receiving a FALSE result, the function will also provide reasoning as to why the data is not in second normal form. Here, the previous examples have shown that no keys of size two exist and therefore the data is not in first normal form. Raising the max_size argument fixes that problem, therefore allowing the data to be in second normal form. Although a three-column key is not ideal in a data set with only eleven variables, but this is a useful feature for very large data sets.

#### third_normal

Description: This function tested a data set for third normal form. The two requirements for this form are being in second normal form and no existing transitive dependencies. Transitive dependencies exist when non-key attributes are dependent on other non-key attributes. All variables need to depend exclusively on the key. Third_normal checks for these dependencies by testing if any non-key subset can completely identify (in the same way a key would) another non-key attribute.

An important part of understanding third normal form is there is the possibility for **coincidental uniqueness**. This occurs when a subset of non-key attributes uniquely identifies another non-key attribute in an inconsequential way. While checks are in place to minimize false positives, the chosen language of "possible transitive dependency" was used. Dependencies flagged should be scrutinized for validity.

```{r}
print(third_normal)
```

Example

```{r}
third_normal(auto)
third_normal(auto, max_size = 3)
```

Similar to the logic of the previous normal form functions, a TRUE result implies at least one key exists and all lower normal forms are satisfied. A FALSE result supplies the reasoning for the data not being in third normal form.

Coincidental Uniqueness Example

```{r}
third_normal1 <- function(data, max_size = 2) {
  
  result <- second_normal(data, max_size = max_size)
  key_results <- find_keys(data, max_size = max_size)
  all_cols <- colnames(data)

  if (grepl("FALSE", result)) {
    return("FALSE: Dataset is NOT in Third Normal Form because it is not in Second Normal Form")
  }
  
  for (i in seq_len(nrow(key_results))) {
    
    key_cols <- unlist(strsplit(key_results$columns[i], ", "))
    non_key_cols <- setdiff(all_cols, key_cols)
    
    if (length(non_key_cols) > 1) {
      
      for (col in non_key_cols) {
        
        other_cols <- setdiff(non_key_cols, col)
        
        for (k in 1:length(other_cols)) {
          
          subsets <- combn(other_cols, k, simplify = FALSE)
          
          for (subset in subsets) {
            
            temp <- data[, c(subset, col), drop = FALSE]
            
            if (!any(duplicated(temp[subset]))) {
              return(paste0(
                "FALSE: Dataset is NOT in Third Normal Form. ",
                "Transitive dependency detected: {",
                paste(subset, collapse = ", "),
                "} -> ", col
              ))
            }
          }
        }
      }
    }
  }
  
  return("TRUE")
}

third_normal1(college)
```

The example shows a transitive dependency that was flagged by an older version of third_normal. The non-key attributes of city and ugds can uniquely identify another non-key attribute instnm. This means that the university name can be completely identified by the city and number of undergrads. This is an inconsequential dependency since number of undergrads is coincidentally unique and not a functional dependency. This function issue that caused this was due to the function testing every possible subset, which not only slowed down the function, but treated all attributes, key and non-key, equally. Updates made to this function significantly reduced the likelihood of a false transitive dependency being flagged.

## Process

To develop this package, we followed a process that moved from conceptual design to functional implementation. We began by identifying the specific gaps in existing R tools related to database normalization and key identification. Once those were clarified and we established what features we wanted to provide, we drafted outlines for each function, including expected inputs, outputs, and logical steps required for functionality. 

A major part of this process involved translating relational database concepts, such as candidate keys and three layers of normal forms, into computational procedures. Plain-English logic was written before any code was written, ensuring that our implementation aligned with theoretical definitions and avoided unnecessary complexity. Testing with the chosen data sets allowed us to verify correctness and refine the functions. Throughout this process, each function was developed independently, but was designed to work together. An example of this can be seen in using is_key() inside find_keys() and using both the normal form checks. Practical considerations such as missing data and data types were also evaluated. These considerations helped guide the development of optional arguments and error messages that clarify to the user why a function fails when expected conditions are not met. Future improvements were noted for the package. These included additional customization options, more extensive testing, and more structural alignment with tidyverse.

## Conclusion

While focusing on creating a package to help users evaluate normal forms and identify keys within relational data sets, the package was developed. Implementing functions that check key validity, search for minimal keys, and assess first and second normal form conditions helped us produce a set of tools that make database concepts easier to apply in everyday data analysis.

While the package currently serves as a functional foundation, there is room for future development. Adding more customization, improving accuracy through additional testing, and expanding capabilities to higher normal forms would enhance the usefulness. Together, these developments demonstrate that the package not only achieves the initial goals of evaluating normal form but also provides a robust foundation for future enhancements. This package shows that normalization concepts can be made more accessible, reproducible, and useful for real data workflows.
