---
title: "Working with NormForm"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with NormForm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(NormForm)
```

# Introduction

Relational databases rely on well-structured data to reduce redundancy, ensure consistency, and support efficient analysis. Unfortunately, many real-world datasets are messy, with unclear keys and violations of normalization principles.

`NormForm` is an R package designed to address this gap by providing tools to:

* Identify candidate keys
* Test whether selected attributes form a valid key
* Evaluate whether a dataset satisfies first, second, and third normal forms

These tools help bridge the gap between theoretical concepts and practical data analysis.

---

# Motivation

When working with real data, determining whether a dataset satisfies normalization conditions often requires manual inspection. Identifying valid keys can also become computationally expensive as the number of variables increases.

`NormForm` simplifies these tasks by automating:

* Key detection
* Key validation
* Normal form assessment

This makes the package especially useful for students, researchers, and analysts working with relational data in R.

---

# Example Data

This vignette uses three example datasets included in the package:

* `college`: 1372 observations, 29 variables, includes a unique identifier (`unitid`)
* `auto`: 398 observations, 11 variables, no obvious unique identifier
* `counties`: 100 observations, 18 variables, includes a unique identifier (`RN`)

```{r}
data(college)
data(auto)
data(counties)
```

---

# Working with Keys

## Checking if a Key Exists

The `is_key()` function tests whether a column or set of columns uniquely identifies each row in a dataset.

```{r}
is_key(college, "unitid")
```

This returns `TRUE` if the column is a valid key.

You can also test combinations of columns:

```{r}
is_key(college, c("city", "stabbr"))
```

---

## Finding Candidate Keys

To automatically search for keys, use `find_keys()`:

```{r}
find_keys(auto, max_size = 2)
```

This function:

* Searches combinations of columns
* Identifies minimal keys
* Excludes redundant combinations

The `max_size` argument controls the maximum number of columns considered in combinations.

---

# Evaluating Normal Forms

## First Normal Form (1NF)

A dataset is in first normal form if:

* All columns contain atomic values
* A valid key exists

```{r}
first_normal(college)
```

If the dataset does not satisfy these conditions, the function returns `FALSE` along with a message indicating the issue.

---

## Second Normal Form (2NF)

A dataset is in second normal form if:

1. It is in first normal form
2. No partial dependencies exist

```{r}
second_normal(college)
```

For composite keys, the function checks whether any subset of the key uniquely identifies the data. If so, a partial dependency exists.

Datasets with single-column keys automatically satisfy second normal form.

---

## Third Normal Form (3NF)

A dataset is in third normal form if:

1. It is in second normal form
2. No transitive dependencies exist

```{r}
third_normal(college)
```

Transitive dependencies occur when non-key attributes depend on other non-key attributes rather than directly on the key.

---

# A Complete Workflow

The following example demonstrates a typical workflow using `NormForm`:

```{r}
# Step 1: Identify candidate keys
keys <- find_keys(auto, max_size = 2)

# Step 2: Validate a candidate key
is_key(auto, keys[[1]])

# Step 3: Evaluate normalization
first_normal(auto)
second_normal(auto)
third_normal(auto)
```

This workflow allows users to move from raw data to a structured assessment of relational integrity.

---

# Conclusion

`NormForm` provides a structured and practical approach to evaluating relational datasets. It allows users to:

* Determine whether columns form valid keys
* Identify minimal keys efficiently
* Evaluate first, second, and third normal form

These tools improve data organization and make database normalization concepts easier to apply in real-world analysis.
